---
title: "Impact of Weather in Relation to Disrupted Flights"
output: html_notebook
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(ranger)
library(rsparse)
library(modelr)
library(infer)

x3colours = c("#41b6c4", "#2c7fb8", "#253494")
x2colours = c("#41b6c4", "#253494")

weather_events <- read_csv("../clean_data/weather_events.csv")

all_flights <- read_csv("../clean_data/all_flight_data.csv")

ewr_disruptions <- all_flights %>% 
  mutate(date = date(time_hour), .after = time_hour) %>% 
  mutate(weather_event_flag = case_when(date %in% weather_events$date ~ "Weather Event",
                                        .default = "Normal Weather")) %>% 
  filter(origin == "EWR")
```

**To what extent does extreme weather coincide with disruption of travel?**

Average number of disrupted flights / weather
```{r}
# length of delays (all flights)
ewr_disruptions %>% 
  drop_na(dep_delay) %>%
  filter(dep_delay <= (mean(dep_delay) + 5*sd(dep_delay))) %>% 
  group_by(weather_event_flag) %>% 
  summarise(mean_delay = mean(dep_delay, na.rm = TRUE), name = "mean_length_of_delay") %>% 
  ggplot(aes(x = name, y = mean_delay, fill = weather_event_flag)) +
  geom_col(position = "dodge", colour = "white") +
  labs(x = "",
       y = "Mean Delay (minutes)\n",
       title = "Length of Delays out of Newark",
       fill = "") +
  theme_minimal() +
  scale_x_discrete(labels = "Delay Length") +
  scale_y_continuous(breaks = seq(0, 25, 4)) +
  scale_fill_manual(values = x2colours) +
  theme(plot.margin = unit(c(0.2, 6, 0.2, 0.2), "cm"))
```

## plotting number of delays/cancellations together

```{r message=FALSE}
# delays
num_delays <- ewr_disruptions %>% 
  filter(delayed_flag == 1) %>% 
  group_by(weather_event_flag, date) %>% 
  summarise(num_delays = n()) %>% 
  summarise(mean_num_delay = mean(num_delays))


# cancellations
## all dates where at least 1 flight cancelled
cancellation_dates <- ewr_disruptions %>% 
  filter(cancelled_flag == 1) %>% 
  distinct(date)
## every date in the year
dates <- tibble(date = seq(ymd("2017-01-01"), ymd("2017-12-31"), 1),
               weather_event_flag =
                 case_when(date %in% weather_events$date ~ "Weather Event",
                                        .default = "Normal Weather"),
               cancellations = 0) 
## tibble of dates with no cancellations
zero_cancellations <- anti_join(dates, cancellation_dates, by = "date")

num_cancellations <- ewr_disruptions %>% 
  filter(cancelled_flag == 1) %>% 
  group_by(weather_event_flag, date) %>% 
  summarise(cancellations = n()) %>% 
  full_join(zero_cancellations,
            by = c("weather_event_flag", "date", "cancellations")) %>% # join so we have all 365 days
  summarise(mean_cancellations = mean(cancellations))

full_join(num_delays, num_cancellations, by = "weather_event_flag") %>% 
  pivot_longer(c(mean_num_delay, mean_cancellations)) %>% 
  ggplot(aes(x = name, y = value, fill = weather_event_flag)) +
  geom_col(position = "dodge", colour = "white") +
  labs(x = "",
       y = "Mean Number of Flights per Day\n",
       title = "Cancellations and Delays out of Newark",
       fill = "") +
  scale_fill_manual(values = x2colours) +
  scale_y_continuous(breaks = seq(0, 100, 20)) +
  scale_x_discrete(labels = c("Number of Cancellations", "Number of Delays")) +
  theme_minimal()
```
```{r}
ewr_disruptions %>% 
  count(weather_event_flag)
```



## Hypothesis test: Are was *significantly* more flight disruption during severe weather events?
H~0~: Flights are **not** more likely to be disrupted during a severe weather event.  
H~1~: Flights are more likely to be disrupted during a severe weather event.   
Level of significance chosen to be 5%.

```{r message=FALSE, warning=FALSE}
ewr_disruptions_mini <- ewr_disruptions %>% 
  select(disrupted_flag, weather_event_flag) %>% 
  mutate(disrupted_flag = as.character(disrupted_flag))

observed_stat <- ewr_disruptions_mini %>% 
  specify(disrupted_flag ~ weather_event_flag, success = "1") %>% 
  calculate(stat = "diff in props", order = c("Weather Event", "Normal Weather"))

null_distribution <- ewr_disruptions_mini %>%
  specify(disrupted_flag ~ weather_event_flag, success = "1") %>% 
  hypothesize(null = "independence") %>% 
  generate(reps = 200, type = "permute") %>%
  calculate(stat = "diff in props", order = c("Weather Event", "Normal Weather"))

null_distribution %>%
  visualise() +
  shade_p_value(obs_stat = observed_stat, direction = "right") +
  theme_minimal()

p_value <- null_distribution %>%
  get_p_value(obs_stat = observed_stat, direction = "right")
p_value
```
With a p-value approaching zero, we are forced to reject the null hypothesis.  Hence, unsurprisingly, flights are more likely to be disrupted during a severe weather event..

Number of delayed flights from EWR / interaction with weather events
```{r message=FALSE, warning=FALSE}
disrupted_plot <- ewr_disruptions %>% 
  filter(disrupted_flag == 1) %>% 
  mutate(date = date(time_hour)) %>% 
  group_by(date) %>% 
  summarise(num_flights = n()) %>% 
  mutate(
    num_delayed_moving_avg = slider::slide_dbl(
      .x = num_flights, 
      .f = ~mean(.x, na.rm = TRUE), 
      .before = 2,
      .after = 2
    ))

weather_event_disrupted_plot <- weather_events %>% 
  left_join(disrupted_plot, by = "date")

# flights_per_day <- ewr_disruptions %>% 
#   mutate(date = date(time_hour)) %>% 
#   group_by(date) %>% 
#   summarise(num_flights = n()) %>% 
#   mutate(
#     num_flights_moving_avg = slider::slide_dbl(
#       .x = num_flights, 
#       .f = ~mean(.x, na.rm = TRUE), 
#       .before = 2,
#       .after = 2
#     ))

ggplot() +
  geom_line(data = disrupted_plot, aes(x = date, y = num_flights), alpha = 0.3, colour = "#2c7fb8") +
  geom_line(data = disrupted_plot, aes(x = date, y = num_delayed_moving_avg), alpha = 1, linewidth = 0.7) +
  #geom_line(data = flights_per_day, aes(x = date, y = num_flights_moving_avg), alpha = 1, linewidth = 0.7) +
  #geom_point(data = weather_event_disrupted_plot, aes(x = date, y = num_flights), colour = "red") +
  geom_point(data = weather_event_disrupted_plot %>% slice_max(num_flights, n = 5),
             aes(x = date, y = num_flights), colour = "red") +
  geom_point(data = weather_event_disrupted_plot %>% slice_min(num_flights, n = 5),
             aes(x = date, y = num_flights), colour = "red") +
  ggtext::geom_richtext(
    data = weather_event_disrupted_plot %>% slice_max(num_flights, n = 5),
    aes(x = date, y = 260, label = event),
    angle = 90, size = 3, fill = "#253494", colour = "white") +
  ggtext::geom_richtext(
    data = weather_event_disrupted_plot %>% slice_min(num_flights, n = 5),
    aes(x = date, y = 170, label = event),
    angle = 90, size = 3, fill = "#41b6c4", colour = "white") +
  labs(title = "Departure Disruption at Newark Airport in 2017",
       subtitle = "With 5-day Rolling Average and Selected Severe Weather Events",
       x = "",
       y = "Number of Flights\n") +
  scale_y_continuous(limits = c(15, 335), breaks = seq(0, 350, 50), minor_breaks = NULL) +
  theme_minimal()
```
Legend: this plot purely used to produce a key for the plot above
```{r}
ggplot() +
  geom_line(aes(x = seq(0, 1, 0.25), y = c(3, 3.2, 3, 2.8, 3)),
            alpha = 0.3, colour = "#2c7fb8") +
  geom_line(aes(x = seq(0, 1, 0.25), y = c(2.5, 2.7, 2.5, 2.3, 2.5)),
            alpha = 1, linewidth = 0.7) +
  geom_point(aes(x = 0.5, y = 2), colour = "red") +
  ggtext::geom_richtext(aes(x = 0.5, y = 1.5, label = "Weather"),
                        size = 3, fill = "#253494", colour = "white") +
  ggtext::geom_richtext(aes(x = 0.5, y = 1, label = "Weather"),
                        size = 3, fill = "#41b6c4", colour = "white") +
  geom_point(aes(x = 15, y = 7), colour = "red") +
  geom_text(aes(x = 1.5, y = seq(1, 3, 0.5),
                label = c("Weather Events with Least Disruption", 
                          "Weather Events with Most Disruption",
                          "Severe Weather Event",
                          "5-day Rolling Average",
                          "Actual Disruption")),
                        size = 3, hjust = 0) +
  scale_y_continuous(breaks = c(-5, 7)) +
  scale_x_continuous(breaks = c(-5, 15)) +
  theme_minimal()

```

# worst weather conditions?
```{r}
mark <- 0.01

all_flights <- ewr_disruptions %>%
  summarise(all_flights = sum(disrupted_flag)/n())%>% 
  pull()

low_temp <-  ewr_disruptions %>% 
  filter(temp <= quantile(temp, mark)) %>% 
  summarise(low_temp = sum(disrupted_flag)/n())%>% 
  pull()

low_dewpoint <- ewr_disruptions %>% 
  filter(dewpoint <= quantile(dewpoint, mark)) %>% 
  summarise(low_dewpoint = sum(disrupted_flag)/n()) %>% 
  pull()

high_wind <- ewr_disruptions %>% 
  filter(wind_speed >= quantile(wind_speed, 1-mark)) %>% 
  summarise(high_wind = sum(disrupted_flag)/n())%>% 
  pull()

high_rain <- ewr_disruptions %>% 
  filter(precip >= quantile(precip, 1-mark)) %>% 
  summarise(high_rain = sum(disrupted_flag)/n())%>% 
  pull()

high_humidity <- ewr_disruptions %>% 
  filter(humidity >= quantile(humidity, 1-mark)) %>% 
  summarise(high_humidity = sum(disrupted_flag)/n())%>% 
  pull()

low_pressure <- ewr_disruptions %>% 
  filter(pressure <= quantile(pressure, mark)) %>% 
  summarise(low_pressure = sum(disrupted_flag)/n())%>% 
  pull()

low_visibility <- ewr_disruptions %>% 
  filter(visib <= quantile(visib, mark)) %>% 
  summarise(low_visibility = sum(disrupted_flag)/n())%>% 
  pull()

high_snow <- ewr_disruptions %>% 
  filter(snow >= quantile(snow, 1-mark)) %>% 
  summarise(snow = sum(disrupted_flag)/n())%>% 
  pull()

weather_effects <- tibble(
  reason = c("All Flights", "Low Temp", "Low Dewpoint", "High Wind", "High Rain",
             "High Humidity", "Low Pressure", "Low Visibility", "High Snow"),
  prop = c(all_flights, low_temp, low_dewpoint, high_wind, high_rain,
           high_humidity, low_pressure, low_visibility, high_snow)
)

weather_effects

ggplot(weather_effects, aes(x = reorder(reason, -prop), y = prop,
                            fill = ifelse(reason == "All Flights", "yes", "no"))) +
  geom_col() +
  geom_abline(slope = 0, intercept = all_flights, colour = "#41b6c4", lty = 2) +
  theme(legend.position = "none") +
  labs(x = "",
       y = "\nProportion of Flights Disrupted",
       title = "Disruption in Most Extreme Weather Conditions") +
  scale_fill_manual(values = c("yes" = "#41b6c4", "no" = "#253494"), guide = "none") +
  scale_y_continuous(labels = scales::percent, breaks = seq(0, 1, 0.1)) +
  theme_minimal() +
  coord_flip()
```

# Finding 1% values
```{r}
ewr_disruptions %>% 
  summarise(
    rain = quantile(precip, 1-mark),
    wind = quantile(wind_speed, 1-mark),
    visibility = quantile(wind_speed, mark),
    snow = quantile(snow, 1-mark)
  )
```





